---
title: "Rmarkdown"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Rmarkdown - 보고서 작성
install.packages("rmarkdown")

|:-----------------------------|:----------------------------------------------------------|
|'##'                          |제목                                                       |
|<>                            |링크                                                       |
|**knit**                      |굵은 글씨                                                  |
|```{r} r code ```             |R코드 작성                                                 |
|```{r pressure, echo=FALSE}```|코드 작성되지 않음(결과만 실행)                            |
|```{r pressure, eval=FALSE}```|코드만 실행                                                |
|'echo=FALSE'                  |=을 문자로 작성하고자 할때 코드로 착각할 수 있으므로 따옴표|

* 한글로 보고서
 1) html형식 :  knit-> open in browser-> 인쇄 (한글로 pdf 복사할 수 있다)  
 2) <!-- 또는 Shift+Ctrl+C(windows) , Shift+CMD+C(macOS)-->
 3) pdf document는 한글X, 무조건 영어만 가능

* library는 모두 맨 위에!
코드 안 돌아가면 코드만 나오고 결과는 나오지 않는다.
-> 유기적으로 무조건 돌아가도록 하기

* 데이터 테이블 작성
:-----  좌측정렬
:-----: 중앙정렬
------: 우측정렬


### Code Chunks  
```{r}
norm <- rnorm(100, mean = 0, sd = 1)
head(norm)
```

### Including Plots  
You can also embed plots, for example:
```{r pressure, echo=FALSE}
plot(pressure)
```  
Note that the `echo = FALSE` parameter was added to the code chunk to prevent **printing** of the R code that generated the plot.

## Including dataframe
```{r}
A <- c("a", "a", "b", "b")
B <- c(5, 10, 15, 20)
dataframe <- data.frame(A, B)
plot(dataframe)
```

## Load the .csv file
Note that the `eval= FALSE` parameter was added to the code chunk to prevent **running** of the R code that generated the plot.
```{r, eval=FALSE}
dataframe <- read.csv("~/Desktop/Code/dataframe.csv")
```

## Load libraries  
```{r}
library(dplyr)
```

## Hiding code chunks  
If you don’t want the code of a particular code chunk to appear in the final document, but still want to show the output (e.g. a plot), then you can include `echo = FALSE` in the code chunk instructions.  
```{r, echo = FALSE}
A <- c("a", "a", "b", "b")
B <- c(5, 10, 15, 20)
dataframe <- data.frame(A, B)
plot(dataframe)
```

Similarly, you might want to create an object, but not include both the code and the output in the final `.html` file. To do this you can use, `include = FALSE`. Be aware though, when making reproducible research it’s often not a good idea to completely hide some part of your analysis:

```{r, include = FALSE}
dataframe <- data.frame(A, B)
plot(dataframe)
```

## Inserting tables  
```{r}
dataframe
```

## Kable() function from knitr package  
```{r}
library(knitr)
kable(dataframe, digits = 2)
```

## Manually creaing tables

| Plant | Temp. | Growth |
|:------|:-----:|-------:|
| A     | 20    | 0.65   |
| B     | 20    | 0.95   |
| C     | 20    | 0.15   |

## Inserting plots  
```{r}
library(ggplot2)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color= drv)) + 
     geom_smooth(mapping = aes(linetype = drv), method = 'loess') +
     geom_point()
```

## Helpful resources
https://ourcodingclub.github.io/tutorials/rmarkdown/



